Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-09-16 summary.Rmd
can now compile without files outside repository

```{r}
import::from(magrittr, `%>%`)
library(plot.matrix)
library(ggplot2)
library(plot.matrix)

@@

* *Case $k \neq l$*. $P^{(kl)}$ and $A^{(kl)}$ represent edge probabilities and 
edges between communities $k$ and $l$. Note that $P^{(kl)} = (P^{(lk)})^\top$.  
By definition, $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. As in the 
$k = l$ case, we note that the singular value decomposition 
$P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top$ is one-dimensional and 
$\lambda^{(kl)} = \sigma_{kl} u^{(kl)}$. (We can also note that the SVD of 
$P^{(lk)} = \sigma_{kl}^2 v^{(kl)} (u^{(kl)})^\top$, i.e., 
$\sigma_{kl} = \sigma_{lk}$, $u^{(kl)} = v^{(lk)}$, and 
$v^{(kl)} = u^{(lk)}$.)  
Now consider the Hermitian dilation 
$M^{(kl)} = 2 \begin{bmatrix} 0 & P^{(kl)} \\ P^{(lk)} & 0 \end{bmatrix}$, 
which is a symmetric $(n_k + n_l) \times (n_k + n_l)$ matrix. It can be shown 
that the eigendecomposition of $M^{(kl)}$ is  
$M^{(kl)} = 
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix} \times 
\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix}^\top$.  
Thus treating $M^{(kl)}$ as the edge probability matrix of a GRDPG, we have 
latent positions in $\mathbb{R}^2$ given by 
$\begin{bmatrix} 
  \sigma_{kl} u^{(kl)} & \sigma_{kl} u^{(kl)} \\ 
  \sigma_{kl} v^{(kl)} & -\sigma_{kl} v^{(kl)} 
\end{bmatrix} = 
\begin{bmatrix} 
  \lambda^{(kl)} & \lambda^{(kl)} \\ 
  \lambda^{(lk)} & -\lambda^{(lk)} 
\end{bmatrix}$.  
Now consider $M' = \begin{bmatrix} 0 & A^{(kl)} \\ A^{(lk)} & 0 \end{bmatrix}$. 
Then $M' = M + E'$ where 
$E' = \begin{bmatrix} 0 & E \\ E^\top & 0 \end{bmatrix}$ and $E$ is the 
$n_k \times n_l$ matrix of independent noise (to generate the Bernoulli entries 
in $A^{(kl)}$.  
Then $M'$ is an adjacency matrix drawn from $M^{(kl)}$, so its adjacency 
spectral embedding, given by 
$\begin{bmatrix} 
  \hat{\lambda}^{(kl)} & \hat{\lambda}^{(kl)} \\ 
  \hat{\lambda}^{(lk)} & -\hat{\lambda}^{(lk)} 
\end{bmatrix}$ 
where each $\hat{\lambda}^{(kl)}$ is defined as in Algorithm 4, approximates 
the latent positions of $M^{(kl)}$ by the rate in Theorem 5 of 
\citet{rubindelanchy2017statistical}.  
In this case, the indefinite orthogonal transformation $W_*$ in the GRDPG 
result \cite{rubindelanchy2017statistical} is of the form 
$\hat{U}^\top \hat{U}$. The eigenvalues of $M$ are distinct since the signature 
for this GRDPG is $(1, 1)$, and $\hat{U}^\top \hat{U}$ is block diagonal, 
resulting in $W_* = I$. Therefore, the adjacency spectral embedding of $M'$ is 
a direct estimation of the specific latent positions outlined for $M^{(kl)}$, 
up to sign flip.

@@

butterfly <- R.matlab::readMat('~/Downloads/Raw_butterfly_network.mat')

@@

```{r}
plot.A <- function(A, margins = c(0, 0, 0, 0)) {
  default.mar <- par()$mar
  par(mar = margins)
  plot(A, border = NA, key = NULL, 
       xlab = '', ylab = '', main = '',
       axis.row = NULL, axis.col = NULL, asp = 1)
  par(mar = default.mar)
}
```

@@

```{r}
# this portion is from the code provided by Sengupta and Chen

setwd('~/Downloads/codes_and_data')
library("igraph")
library("clue")

##### Data input and pre-processing #####
foo=read.table("BritishMPcomm.txt",header=T)
mat=read.table("politicsuk-retweets.mtx",header=F)
mat = mat[-1,]
A = matrix(0,nrow(foo),nrow(foo))
for (k in 1:nrow(mat)){ #adjacency matrix with 5 parties
  edgei = mat[k,1]
  edgej = mat[k,2]
  i = which(foo[,1]==edgei)
  j = which(foo[,1]==edgej)
  A[i,j]<-1;A[j,i]<-1
}
b.true <- foo[,2]
levels(b.true) <- 1:5
b.true <- as.numeric(b.true)
nodes1 = which(b.true==1); nodes2 = which(b.true==2)
nodes = c(nodes1, nodes2)
A <- A[nodes,nodes] # Select only the two big communities
G<-graph.adjacency(A, mode="undirected")
foo1 <- clusters(G, mode="strong") 
nodes2 <- which(foo1$membership==1) # largest connected component (basically 31 nodes are disconnected)
A <- A[nodes2,nodes2] # form network with 329 nodes
b.true <- b.true[nodes2]
```

```{r}
zhat <- cluster.pabm(A, 2, use.all = TRUE, normalize = FALSE)
clust.acc <- mean(zhat == b.true)
```

```{r mp, cache = TRUE, fig.width = 2, fig.height = 2, fig.cap = 'Adjacency matrix of the British MPs dataset after sorting by the clustering outputted by Algorithm 2.'}
plot.A(A[order(zhat), order(zhat)])
```

In the second example, we applied Algorithm 2 to the British MPs Twitter 
network \cite{greene2013producing}. The original data consists of 419 vertices, 
each corresponding to a British MP belonging to one of five political parties. 
For this data analysis, we subsetted the data as described by 
\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the same 
dataset, resulting in 329 vertices and two parties ($n = 329, K = 2$). 
\citet{307cbeb9b1be48299388437423d94bf1} were able to obtain a classification 
accuracy of `r round(328 / 329 * 100, 1)`\% via modularity maximization. 
Algorithm 2 obtained a lower 
classification accuracy of 99.1\%. We can also note 
that Algorithm 2 had a much shorter runtime than modularity maximization.

>>>>

```{r}
import::from(magrittr, `%>%`)
library(plot.matrix)
library(ggplot2)

@@

* *Case $k \neq l$*. $P^{(kl)}$ and $A^{(kl)}$ represent edge probabilities and 
edges between communities $k$ and $l$. Note that $P^{(kl)} = (P^{(lk)})^\top$.  
By definition, $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. As in the 
$k = l$ case, we note that the singular value decomposition 
$P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top$ is one-dimensional and 
$\lambda^{(kl)} = \sigma_{kl} u^{(kl)}$. (We can also note that the SVD of 
$P^{(lk)} = \sigma_{kl}^2 v^{(kl)} (u^{(kl)})^\top$, i.e., 
$\sigma_{kl} = \sigma_{lk}$, $u^{(kl)} = v^{(lk)}$, and 
$v^{(kl)} = u^{(lk)}$.)  
Now consider the Hermitian dilation 
$M^{(kl)} = 2 \begin{bmatrix} 0 & P^{(kl)} \\ P^{(lk)} & 0 \end{bmatrix}$, 
which is a symmetric $(n_k + n_l) \times (n_k + n_l)$ matrix. It can be shown 
that the eigendecomposition of $M^{(kl)}$ is  
$M^{(kl)} = 
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix} \times 
\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
\begin{bmatrix} u^{(kl)} & -u^{(kl)} \\ v^{(kl)} & v^{(kl)} \end{bmatrix}^\top$.  
Thus treating $M^{(kl)}$ as the edge probability matrix of a GRDPG, we have 
latent positions in $\mathbb{R}^2$ given by 
$\begin{bmatrix} 
  \sigma_{kl} u^{(kl)} & \sigma_{kl} u^{(kl)} \\ 
  \sigma_{kl} v^{(kl)} & -\sigma_{kl} v^{(kl)} 
\end{bmatrix} = 
\begin{bmatrix} 
  \lambda^{(kl)} & \lambda^{(kl)} \\ 
  \lambda^{(lk)} & -\lambda^{(lk)} 
\end{bmatrix}$.  
Now consider $\hat{M}^{(kl)} = \begin{bmatrix} 0 & A^{(kl)} \\ A^{(lk)} & 0 \end{bmatrix}$. 
Then $\hat{M}^{(kl)} = M^{(kl)} + E'$ where 
$E' = \begin{bmatrix} 0 & E \\ E^\top & 0 \end{bmatrix}$ and $E$ is the 
$n_k \times n_l$ matrix of independent noise (to generate the Bernoulli entries 
in $A^{(kl)}$.  
Then $\hat{M}^{(kl)}$ is an adjacency matrix drawn from $M^{(kl)}$, so its adjacency 
spectral embedding, given by 
$\begin{bmatrix} 
  \hat{\lambda}^{(kl)} & \hat{\lambda}^{(kl)} \\ 
  \hat{\lambda}^{(lk)} & -\hat{\lambda}^{(lk)} 
\end{bmatrix}$ 
where each $\hat{\lambda}^{(kl)}$ is defined as in Algorithm 4, approximates 
the latent positions of $M^{(kl)}$ by the rate in Theorem 5 of 
\citet{rubindelanchy2017statistical}.  
In this case, the indefinite orthogonal transformation $W_*$ in the GRDPG 
result \cite{rubindelanchy2017statistical} is of the form 
$U^\top \hat{U}$. The eigenvalues of $M$ are distinct since the signature 
for this GRDPG is $(1, 1)$, and $U^\top \hat{U}$ is block diagonal, 
resulting in $W_* \stackrel{P}{\to} I$. Therefore, the adjacency spectral embedding of $\hat{M}^{(kl)}$ is 
a direct estimation of the specific latent positions outlined for $M^{(kl)}$, 
up to sign flip.

@@

butterfly <- R.matlab::readMat('data/Raw_butterfly_network.mat')

@@

```{r}
# plot.A <- function(A, margins = c(0, 0, 0, 0)) {
#   default.mar <- par()$mar
#   par(mar = margins)
#   plot(A, border = NA, key = NULL, 
#        xlab = '', ylab = '', main = '',
#        axis.row = NULL, axis.col = NULL, asp = 1)
#   par(mar = default.mar)
# }

# plot.A <- function(A, labels) {
#   heatmap(A[order(labels), order(labels)],
#           margins = c(0, 0),
#           revC = TRUE)
# }

plot.A <- function(A, labels) {
  # https://www.r-graph-gallery.com/79-levelplot-with-ggplot2.html
  A[order(labels), order(labels)] %>% 
    tibble::as_tibble() %>% 
    tibble::rowid_to_column(var = 'X') %>% 
    tidyr::gather(key = 'Y', value = 'Z', -1) %>% 
    dplyr::mutate(Y = as.numeric(gsub('V', '', Y))) %>% 
    ggplot() + 
    geom_tile(aes(x = X, y = Y, fill = Z)) + 
    coord_fixed() + 
    labs(x = NULL, y = NULL) + 
    theme_void() + 
    theme(legend.position = 'none',
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank()) + 
    scale_x_continuous(expand = c(0, 0)) + 
    scale_y_continuous(expand = c(0, 0)) + 
    scale_fill_gradient(low = 'yellow', high = 'red')
}
```

@@

```{r}
# this portion is from the code provided by Sengupta and Chen
# the data were generated by scripts from S&C, then saved as RDS files

bmp <- readRDS('data/bmp.rds')
A.bmp <- bmp$A
z.bmp <- bmp$z

pb <- readRDS('data/pb.rds')
A.pb <- pb$A
z.pb <- pb$z

dblp <- readRDS('data/dblp.rds')
A.dblp <- dblp$A
z.dblp <- dblp$z
```

```{r, cache = TRUE}
zhat.bmp <- cluster.pabm(A.bmp, 2, use.all = TRUE, normalize = TRUE)
bmp.acc <- mean(zhat.bmp == z.bmp)

zhat.pb <- cluster.pabm(A.pb, 2, use.all = TRUE, normalize = TRUE, 
                        p = 2, q = 0)
pb.acc <- mean(zhat.pb == z.pb)

zhat.dblp <- cluster.pabm(A.dblp, 2, use.all = TRUE, normalize = FALSE)
dblp.acc <- 1 - mean(zhat.dblp == z.dblp)
```

In the second example, we applied Algorithm 2 to the British MPs Twitter 
network \cite{greene2013producing}, the Political Blogs network and the DBLP network. For this data analysis, we subsetted the data as described by 
\citet{307cbeb9b1be48299388437423d94bf1} for their analysis of the same 
networks. Our methods underperformed compared to modularity maximization, although performance is comparable for the British MPs and DBLP networks.

```{r}
dplyr::tibble(network = c('British MPs', 'Political blogs', 'DBLP'),
              vertices = c(nrow(A.bmp), nrow(A.pb), nrow(A.dblp)),
              `error (modularity maximization)` = c(round(1 / 329, 3),
                                                    round(61 / 1222, 3),
                                                    round(62 / 2203, 3)),
              `error (algorithm 2)` = round(1 - c(bmp.acc, 
                                                  pb.acc,
                                                  dblp.acc), 3)) %>% 
  knitr::kable(caption = paste('Community detection error rates', 
                               'for modularity maximization',
                               '(\\citet{307cbeb9b1be48299388437423d94bf1})',
                               'and Algorithm 2.'))
```

```{r mp, cache = TRUE, fig.width = 8, fig.height = 2, fig.cap = 'Adjacency matrices of (from left to right) the British MPs, Political Blogs, and DBLP networks after sorting by the clustering outputted by Algorithm 2.'}
gridExtra::grid.arrange(plot.A(A.bmp, zhat.bmp),
                        plot.A(A.pb, zhat.pb),
                        plot.A(A.dblp, zhat.dblp),
                        ncol = 3)

```
